### 2014-03-11 docs/article.rmd
mer

```{r}
"
Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.

@@

## Examples of Undesired consequences
As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness. The first is a "worst-case"scenario where a few batches are completly missing groups coupled with a a priori assumption of no group difference (24421904). The second compares different cell types and is only mildly unbalanced ("data set 2" from 16632515). We tested their most highlighted group difference for differentially expressed genes using the described approach with ComBat and an alternative approach handling batch inside the test. In addition we performed one sanity check permuting the group labels whitin batches, and one sanity check using random numbers. The different p-value distributions are shown in Figure 2.

![fig. 1](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)

```{r}
"
3. Vise at det er variasjon i stoerrelsen paa feilen introdusert av ComBat
ved a)worst-case scenario (MS-artikel)og b) mild-case (data set 2 fra
ComBat). Vises med p-verdi plot med/uten ComBat og eventuelt antall
signifikante prober.
"
```
## Discussion

## Sliding application area of ComBat
In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only examplified in the supplementary information. However, over the years this feature became more important judging from advice given by the author to users
(
[1](https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ)
,[2](https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ)
,[3](https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ)
,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
).
And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists). Thus, this problematic use has likely been common.

## Motivation for this Warning
Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects. Upon realizing that the confidence on our group differences were exagerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recomended our usage (16632515, 22257669). In addition, other works looking into the problem of batch effects were mostly recomending ComBat without much concern (22682473, 22133085). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070). A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070). Often no effort was done in order to substantiate the existance of batch effects in their data, ecxept for stating the presence of batches (24391845?, 18414638, 21731603, 23630272). In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref). The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
Taken together we fear that many published results from data adjusted by ComBat are completly or partially false. And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficently unbalanced design will continue to be used.

```{r}
"

>>>>

```{r echo=FALSE}
"
Kort oppsummere behovet for batch-korrigering: generelt og i microarraydata.

@@

## Examples of Undesired consequences
As the amount of false positive results as a consequence of trying to retain group differences depends on the group/batch balance, we will show two examples with varying degree of unbalancedness.

1) In the first experiment(24421904), cells are treated glatiramer acetate (a medicine for multiple sclerosis) or a generic and mRNA was measured on microarrays alongside control samples. A batch effect correlated to the array platform (Illumina WG-6\_V2) was observed and adjusted for with ComBat whereafter the data was tested for differetially expressed genes, yielding a list of 1000 genes (Table S5 24421904). Unfortunately the treatment/batch design was unbalanced with several batches having only one of the main treatment of interest. When we re-analyzed their data without using ComBat, but instead blocked for batch effect in limma, only 9 genes were found (FDR<0.05). Additional sanity checks were also applied and the distribution of p-values for differents settings are shown in Fig. 2.  Our conclusion is that most of the genes reported as differtially expresssed in (24421904) are false positives. This example is a sort of "worst case" scenario for applying ComBat, since it both has a very unbalanced batch/group design and a a priori assumpotion of no difference. The R-code for our analysis and a more extensive report can be downloaded from github (https://github.com/vegardny/combat_tests.git).

![fig. 2](../reanalysis/Fadi_Towfic_generic_medicine/figure/pvaluestowfic.svg)
**Figure 3**.
P-value distribution for the main comparison in (data set 2 24421904) glatiramer acetate vs. generic.  
- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
- d) 10 runs of real data with the glatiramer acetate("DP") and generic("N") labels permuted within batches, followed by ComBat adjustment and significance test in limma.

2) The second example is taken from the supporting information for the original ComBat article (16632515) where it is denoted "data set 2". Cells inhibited for the expression of the TAL1 gene were compared to controls on a microarray platform. The experiment was conducted on three different time points (used as batches) with a total of 30 samples with a fairly balanced treatment/control across batches (6/2, 3/4 and 9/6). ComBat was applied followed by a t-test identifying differentially expressed genes. First, we reproduced their analysis including the adjustment by ComBat, but using limma instead of the t-test, resulting in 1003 probes (q<0.05).  Then, we analysed their data blocking for batch in limma and not using ComBat, resulting in 377 probes (q<0.05). In addition the two sanity checks outline above were performed. Comparing the P-value distributions for the alternative analyses (Fig.3 a,b) does not indicate a huge difference, in contrast to results above for(24421904).

![fig. 3](../reanalysis/ComBat_original/figure/pvaluesjohnson.svg)  
**Figure 3**.
P-value distribution for the main comparison in (data set 2 16632515) for  
- a) Real data adjusted by ComBat as described in the article followed by a significance test in limma. 
- b) Real data not adjusted by ComBat, tested for significance in limma using batch as a blocking factor. 
- c) Real data is substituted with random numbers from a normal distribution (mean=0, std=1), but the batch/group design is retained, followed by batch adjustment in ComBat and a significance test in limma.
- d) 10 runs of real data with the group labels permuted within batches, followed by ComBat adjustment and significance test in limma.

## Discussion

## Sliding application area of ComBat
In the original ComBat article(2007) it is clear that the primary motivation behind ComBat was to employ an Emprical Bayes method for batch-effect removal. The feature of retaining group differences for unbalanced designs is optional and seems to be subordinate, only exemplified in the supplementary information. However, over the years this feature became more important judging from advice given by the author to users
(
[1](https://groups.google.com/d/msg/combat-user-forum/eSVSKwGtuyE/ZIWV2juYmmAJ)
,[2](https://groups.google.com/d/msg/combat-user-forum/Vkb9p7wekd4/h5Etie7FTVQJ)
,[3](https://groups.google.com/d/msg/combat-user-forum/fpBTcgDjiR8/uo4QIZL4sZgJ)
,[4](https://groups.google.com/d/msg/combat-user-forum/26FZlgU2LFQ/W6U_Lhh_64EJ)
).
And when ComBat was incorporated in the sva package ( [22257669](http://bioinformatics.oxfordjournals.org/cgi/pmidlookup?view=long&pmid=22257669), 2012), inclusion of group labels was made almost mandatory (an undocumented option of passing a NULL value exists). Thus, this problematic use has likely been common.

## Motivation for this Warning
Our knowledge of ComBat came through a typical use case when trying to salvage unbalanced data which had batch effects. Upon realizing that the confidence on our group differences were exaggerated, the literature was searched for a better understanding of correct use and potential overseen limitations of ComBat. But the authors of ComBat and the sva package recommended our usage (16632515, 22257669). In addition, other works looking into the problem of batch effects were mostly recommending ComBat without much concern (22682473, 22133085). A brief inquiry into  some of the articles citing ComBat (574 Google Scholar) revealed few problems, and their method descriptions regarding ComBat were mostly sparse, limited to one or two sentences (24391845, 18414638, 21731603,23630272, 24584070). A further indication of their carefree use of this potentially devastating procedure was the frequent neglect to state the program parameters, i.e batch-labels (18414638, 21731603) or group labels (24391845, 18414638, 21731603, 23630272, 23482648, 24584070). Often no effort was done in order to substantiate the existence of batch effects in their data, except for stating the presence of batches (24391845?, 18414638, 21731603, 23630272). In one instance ComBat was even applied on data were effects due to batch were investigated, but not found (husker ikke ref). The incorporation of the method into analysis pipelines (16642009,TCGA) and other packages (23452776, 21937664) could make its usage even more trivial and parameters setting harder to perceive.
Taken together we fear that many published results from data adjusted by ComBat are completely or partially false. And knowing that scientist don't give up when one analysis fails, a method that almost ensures a result given a sufficiently unbalanced design will continue to be used.

```{r echo=FALSE}
"
